An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-21 article/latex_files/ch2/2-1-implicit-models.tex
fixed spaces and commas

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterised
family of probability density functions
\(\{ p(\yb|\thetab) \}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. Practically, in these scenarios, we
can only a access a simulator \( M_r(\thetab) \), i.e. a random
black-box machine (computer code) that generates samples \(\yb\) in a
stochastic manner for any given a set of parameters \(\thetab\);
\( M_r(\thetab) \rightarrow \yb \). We can isolate the randomness by
introducing the stochastic nuisance variables \(\ub \sim p(\ub)\) so
that the simulator becomes a deterministic mapping \(\simulator\) that
maps \((\thetab, \ub)\) to the data \(\yb\), i.e.\
\(\yb=\simulator(\thetab,\ub)\). Within the computer code, the
distribution \(p(\ub)\) is defined as the random number generating
process.

Simulator-based models provide considerable modelling freedom; any
physical process that can be conceptualised as a computer program of
finite steps can be modelled as a simulator-based model without any
compromise. The modelling freedom allows for any amount of hidden
(unobserved) internal variables or rule-based decisions. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as e.g.~genetics, epidemiology, or neuroscience. 
Further background on simulator-based models and
example applications can be found in the articles
by \citet{Gutmann2016, Lintusaari2017, Sisson2018, Cranmer2020}.

The modelling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting the
observed data as \(\data\), the main difficulty is that the likelihood
function \(L(\thetab) = p(\data|\thetab)\) is generally
intractable. To better see the sources of the intractability, and to
address them, we go back to the basic characterisation of the
likelihood as the (rescaled) probability of a parameter of the model
to generate data \(\yb\) that is similar to the observed data
\(\data\). More formally, the likelihood \(L(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e.\ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterisation of the likelihood in \eqref{eq:likelihood}
highlights two sources of intractability: The first is the computation
of the probability
\(\Pr(\simulator(\thetab,\ub) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\).  Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(L_{d, \epsilon}(\thetab)\)
\begin{equation} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

@@

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto L_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

The approximation in \eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractabilities of the likelihood
function in \eqref{eq:likelihood}. Other strategies include modelling
the (stochastic) relationship between \(\thetab\) and \(\yb\), and its
reverse, or framing likelihood-free inference as a ratio estimation
problem, see for example \citet{Blum2010, Wood2010, Papamakarios2016,
  Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both
OMC and robust OMC, which we introduce next, are based on the
approximation in \eqref{eq:approx_likelihood}.

>>>>

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterised
family of probability density functions
\(\{ p(\yb|\thetab) \}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we
can only a access a simulator \( m_r(\thetab) \), i.e., a random
black-box machine (computer code) that generates samples \(\yb\) in a
stochastic manner for any given a set of parameters \(\thetab\);
\( m_r(\thetab) \rightarrow \yb \). We can isolate the randomness by
introducing the stochastic nuisance variables \(\ub \sim p(\ub)\) so
that the simulator becomes a deterministic mapping \(\simulator\) that
maps \((\thetab, \ub)\) to the data \(\yb\), i.e., \
\(\yb=\simulator(\thetab,\ub)\). Within the computer code, the
distribution \(p(\ub)\) is defined as the random number generating
process.

Simulator-based models provide considerable modelling freedom; any
physical process that can be conceptualised as a computer program of
finite steps can be modelled as a simulator-based model without any
compromise. The modelling freedom allows for any amount of hidden
(unobserved) internal variables or rule-based decisions. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as, e.g., genetics, epidemiology or
neuroscience. Further background on simulator-based models and
example applications can be found in the articles
by~\citet{gutmann2016, lintusaari2017, sisson2018, cranmer2020}.

The modelling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting the
observed data as \(\data\), the main difficulty is that the likelihood
function \(l(\thetab) = p(\data|\thetab)\) is generally
intractable. To better see the sources of the intractability, and to
address them, we go back to the basic characterisation of the
likelihood as the (rescaled) probability of a parameter of the model
to generate data \(\yb\) that is similar to the observed data
\(\data\). More formally, the likelihood \(l(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  l(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in b_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e., \ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterisation of the likelihood in~\eqref{eq:likelihood}
highlights two sources of intractability; the first is the computation
of the probability
\(\Pr(\simulator(\thetab,\ub) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\). Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(l_{d, \epsilon}(\thetab)\)
\begin{equation} \label{eq:approx_likelihood}
  l_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

@@

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto l_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

The approximation in~\eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractabilities of the likelihood
function in~\eqref{eq:likelihood}. Other strategies include modelling
the (stochastic) relationship between \(\thetab\) and \(\yb\), and its
reverse, or framing likelihood-free inference as a ratio estimation
problem, see for example \citet{blum2010, Wood2006, Papamakarios2016,
  Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both
OMC and robust OMC, which we introduce next, are based on the
approximation in~\eqref{eq:approx_likelihood}.

### 2022-12-21 article/latex_files/ch2/2-2-OMC.tex
fixed spaces and commas

Our description of OMC \citet{Meeds2015} follows \citet{Ikonomov2019}
who base their explanation of OMC on the approximate likelihood
function in \eqref{eq:approx_likelihood}. With the indicator function
%
\begin{equation} \label{eq:indicator} \indicator{\region(\yb)}(\xb)=
  \left\{
    \begin{array}{ll}
      1 & \mbox{if } \xb \in \region(\yb) \\
      0 & \mbox{otherwise} 
    \end{array} \right. \end{equation}
%
we can write the approximate likelihood function \(L_{d, \epsilon}(\thetab)\) in \eqref{eq:approx_likelihood} as
\begin{gather} \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data)) =
  \int_{\yb \in B_{d,\epsilon}(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb
\end{gather}
which can be approximated as a sample's average
\begin{equation}
L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\region (\data)} (\yb_i)
 \quad \text{ where } \yb_i = \simulator(\thetab, \ub_i),\; \ub_i \sim p(\ub). \label{eq:alt_view}
\end{equation}
We thus approximate the likelihood of a specific \(\thetab\) by
counting the fraction of samples that lie inside a volume around the
observations. Isolating the randomness of the simulator via the
\(\ub_i\) has two crucial consequences: First, we can sample the
nuisance variables \(\ub_i \sim p(\ub)\) only once and reuse them to
approximate \(L_{d,\epsilon}(\thetab)\) for different
\(\thetab\). Second, since \(\yb_i=\simulator(\thetab, \ub_i)\) is a
function of \(\thetab\), checking whether \(\yb_i\) is contained in
\(\region (\data)\) is the same as checking whether \(\thetab\) is in
the acceptance region
\(\accregioni = \{ \thetab: \simulator(\thetab, \ub_i) \in
\region(\data) \}\). This leads to the likelihood approximation
\begin{equation}
L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\accregioni}(\thetab) \label{eq:alt_view_theta}
\end{equation}
and the corresponding approximate posterior
\begin{equation} \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data) \propto
  p(\thetab) \sum_i^N  \indicator{\accregioni}(\thetab).
\end{equation}
As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterising the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realisation \(\ub_i \sim p(\ub)\), i.e.\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

>>>>

Our description of OMC \citet{Meeds2015} follows \citet{Ikonomov2019}
who base their explanation of OMC on the approximate likelihood
function in~\eqref{eq:approx_likelihood}. With the indicator function
%
\begin{equation} \label{eq:indicator} \indicator{\region(\yb)}(\xb)=
  \left\{
    \begin{array}{ll}
      1 & \mbox{if } \xb \in \region(\yb) \\
      0 & \mbox{otherwise} 
    \end{array} \right. \end{equation}
%
we can write the approximate likelihood function \(L_{d, \epsilon}(\thetab)\) in \eqref{eq:approx_likelihood} as
\begin{gather} \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data)) =
  \int_{\yb \in B_{d,\epsilon}(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb
\end{gather}
which can be approximated as a sample's average
\begin{equation}
L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\region (\data)} (\yb_i)
 \quad \text{ where } \yb_i = \simulator(\thetab, \ub_i),\; \ub_i \sim p(\ub). \label{eq:alt_view}
\end{equation}
We thus approximate the likelihood of a specific \(\thetab\) by
counting the fraction of samples that lie inside a volume around the
observations. Isolating the randomness of the simulator via the
\(\ub_i\) has two crucial consequences: First, we can sample the
nuisance variables \(\ub_i \sim p(\ub)\) only once and reuse them to
approximate \(L_{d,\epsilon}(\thetab)\) for different
\(\thetab\). Second, since \(\yb_i=\simulator(\thetab, \ub_i)\) is a
function of \(\thetab\), checking whether \(\yb_i\) is contained in
\(\region (\data)\) is the same as checking whether \(\thetab\) is in
the acceptance region
\(\accregioni = \{ \thetab: \simulator(\thetab, \ub_i) \in
\region(\data) \}\). This leads to the likelihood approximation
\begin{equation}
L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\accregioni}(\thetab) \label{eq:alt_view_theta}
\end{equation}
and the corresponding approximate posterior
\begin{equation} \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data) \propto
  p(\thetab) \sum_i^N  \indicator{\accregioni}(\thetab).
\end{equation}
As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterising the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realisation \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

### 2022-12-21 article/latex_files/ch3/3-0-design-principles.tex
final version
fixed spaces and commas

If \(d_i(\thetab)\) is differentiable, using a gradient-based method
is advised for obtaining \(\thetab_i^*\) faster. In this case, the
gradients \(\nabla_{\thetab} d_i\) gradients are approximated
automatically with finite-differences. This approximation requires two
evaluations of \(d_i\) for \emph{each} parameter
\(\theta_m, m \in \{1, \ldots, D\}\)\footnote{\(D\) is the
  dimensionality of \(\thetab\), i.e.  \(\thetab \in \mathbb{R}^D\)},
which works in low-dimensional problems. If \(d_i(\thetab)\) is not
differentiable, Bayesian Optimisation can be used as an alternative
solution. In this scenario, the training part becomes slower due to
fitting of the surrogate model and the blind optimisation steps.

@@

\begin{algorithm}[!ht]
	\caption{Approximation \(\accregioni\) with a bounding box \(\hat{\accregioni}\); 
  Requires: a model of distance \(d_i(\thetab)\), 
  an optimal point \(\thetab_i^*\), 
  a number of refinements \(K\), 
  a step size \(\eta\_\text{start}\), 
  maximum iterations \(M\) and 
  a curvature matrix \(\hessian_i\) (\(\jac_i^T\jac_i \) or GP Hessian)}\label{alg:region_construction}
	\begin{algorithmic}[1]
	\State Compute eigenvectors \(\vb_{m}\) of \(\hess_i\) {\scriptsize (\(d = 1,\ldots,D)\)}
	\For{\(m \gets 1 \textrm{ to } D\)}
		\State \(\Tilde{\thetab} \gets \thetab_i^*\) \label{algstep:box_constr_start}
		\State \(k \gets 0\)
		\State \(\eta \gets \eta\_\text{start}\) \Comment{Initialize \(\eta\)}
		\Repeat
          \State \(j \gets 0\)
        	\Repeat
            \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \eta \ \vb_{m}\) \Comment{Large step size \(\eta\).}
            \State \(j \gets j + 1\)
        	\Until{\(d(g(\Tilde{\thetab}, \ub=\ub_i), \data) > \epsilon\) or \(j \geq M\)} \Comment{Check distance or maximum iterations}
        	\State \(\Tilde{\thetab} \gets \Tilde{\thetab} - \eta \ \vb_{m}\)
        	\State \(\eta \gets \eta/2\) \Comment{More accurate region boundary}
        	\State \(k \gets k + 1\)
      \Until \(k = K\)
      \If{\(\Tilde{\thetab} = \thetab_i^*\)} \Comment{Check if no step has been done}
        \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \dfrac{\eta\_{\text{start}}}{2^K} \ub_m\) \Comment{Then, make the minimum step}
      \EndIf \label{algstep:box_constr_end}
    	\State Set \(\Tilde{\thetab}\) as the positive end point along \(\vb_m\)
    	\State Runc steps~\ref{algstep:box_constr_start}~-~\ref{algstep:box_constr_end} for \(\vb_m = - \vb_m\) and set \(\Tilde{\thetab}\) as the negative end point along \(\mathbf{v}_{m}\)
	\EndFor
	\State Fit a rectangular box around the region end points and define \(q_i\) as uniform distribution
	\end{algorithmic}
\end{algorithm}

>>>>

If \(d_i(\thetab)\) is differentiable, using a gradient-based method
is advised for obtaining \(\thetab_i^*\) faster. In this case, the
gradients \(\nabla_{\thetab} d_i\) gradients are approximated
automatically with finite-differences. This approximation requires two
evaluations of \(d_i\) for \emph{each} parameter
\(\theta_m, m \in \{1, \ldots, D\}\)\footnote{\(D\) is the
  dimensionality of \(\thetab\), i.e.,  \(\thetab \in \mathbb{R}^D\)},
which works in low-dimensional problems. If \(d_i(\thetab)\) is not
differentiable, Bayesian Optimisation can be used as an alternative
solution. In this scenario, the training part becomes slower due to
fitting of the surrogate model and the blind optimisation steps.

@@

\begin{algorithm}[!ht]
	\caption{Approximation \(\accregioni\) with a bounding box \(\hat{\accregioni}\); 
  Requires: a model of distance \(d_i(\thetab)\), 
  an optimal point \(\thetab_i^*\), 
  a number of refinements \(K\), 
  a step size \(\eta\_\text{start}\), 
  maximum iterations \(M\) and 
  a curvature matrix \(\hessian_i\) (\(\jac_i^T\jac_i \) or GP Hessian)}\label{alg:region_construction}
	\begin{algorithmic}[1]
	\State Compute eigenvectors \(\vb_{m}\) of \(\hess_i\) {\scriptsize (\(d = 1,\ldots,D)\)}
	\For{\(m \gets 1 \textrm{ to } D\)}
		\State \(\Tilde{\thetab} \gets \thetab_i^*\) \label{algstep:box_constr_start}
		\State \(k \gets 0\)
		\State \(\eta \gets \eta\_\text{start}\) \Comment{Initialize \(\eta\)}
		\Repeat
          \State \(j \gets 0\)
        	\Repeat
            \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \eta \ \vb_{m}\) \Comment{Large step size \(\eta\).}
            \State \(j \gets j + 1\)
        	\Until{\(d(g(\Tilde{\thetab}, \ub=\ub_i), \data) > \epsilon\) or \(j \geq M\)} \Comment{Check distance or maximum iterations}
        	\State \(\Tilde{\thetab} \gets \Tilde{\thetab} - \eta \ \vb_{m}\)
        	\State \(\eta \gets \eta/2\) \Comment{More accurate region boundary}
        	\State \(k \gets k + 1\)
      \Until \(k = K\)
      \If{\(\Tilde{\thetab} = \thetab_i^*\)} \Comment{Check if no step has been done}
        \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \dfrac{\eta\_{\text{start}}}{2^K} \vb_m\) \Comment{Then, make the minimum step}
      \EndIf \label{algstep:box_constr_end}
    	\State Set \(\Tilde{\thetab}\) as the positive end point along \(\vb_m\)
    	\State Runc steps~\ref{algstep:box_constr_start}~-~\ref{algstep:box_constr_end} for \(\vb_m = - \vb_m\) and set \(\Tilde{\thetab}\) as the negative end point along \(\mathbf{v}_{m}\)
	\EndFor
	\State Fit a rectangular box around the region end points and define \(q_i\) as uniform distribution
	\end{algorithmic}
\end{algorithm}

### 2022-12-21 article/latex_files/ch3/3-1-implementation-principles.tex
fixed spaces and commas

The overview of our implementation is illustrated in
Figure~\ref{fig:romc_overview}. Following \proglang{Python}'s naming
principles, the methods starting with an underscore (green rectangles)
represent internal (private) functions, whereas the rest (blue
rectangles) are the methods exposed as the public API. In
Figure~\ref{fig:romc_overview}, it can be easily observed that the
implementation follows Algorithm~\ref{alg:romc_algorithm}. The
training part includes all the steps until the computation of the
proposal regions i.e.~sampling the nuisance variables, defining the
optimisation problems, solving them, constructing the regions and
fitting local surrogate models. The inference part comprises of
evaluating the unnormalised posterior (and the normalised when is
possible), sampling and computing an expectation. We also provide some
utilities for inspecting the training process, such as plotting the
histogram of the final distances or visualising the constructed
bounding boxes. Finally, in the evaluation part, we provide two
methods for evaluating the inference; (a) computing the Effective
Sample Size (ESS) of the samples and (b) measuring the divergence
between the approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method
  (i.e.\ ABC Rejection Sampling), and we would like to measure the
  discrepancy between the two approximations.}

@@

\begin{Code}
---------------------------------- python ----------------------------------
>>> elfi.ROMC(<output_node>, parallelize=True)
----------------------------------------------------------------------------
\end{Code}

@@

In the following code snippet, we code the model at \pkg{ELFI}.

\begin{Code}
------------------------------ python snippet ------------------------------
  import elfi
  import scipy.stats as ss
  import numpy as np

  def simulator(t1, batch_size=1,random_state=None):
    c = 0.5 - 0.5**4
    if t1 < -0.5:
        y = ss.norm(loc=-t1-c, scale=1).rvs(random_state=random_state)
    elif t1 <= 0.5:
        y = ss.norm(loc=t1**4, scale=1).rvs(random_state=random_state)
    else:
        y = ss.norm(loc=t1-c, scale=1).rvs(random_state=random_state)
    return y

@@

  # Elfi graph
  t1 = elfi.Prior('uniform', -2.5, 5)
  sim = elfi.Simulator(simulator, t1, observed=y)
  d = elfi.Distance('euclidean', sim)

  # Initialise the ROMC inference method
  bounds = [(-2.5, 2.5)] # limits of the prior
  parallelize = True # activate parallel execution
  romc = elfi.ROMC(d, bounds=bounds, parallelize=parallelize)
----------------------------------------------------------------------------
\end{Code}

>>>>

The overview of our implementation is illustrated in
Figure~\ref{fig:romc_overview}. Following \proglang{Python}'s naming
principles, the methods starting with an underscore (green rectangles)
represent internal (private) functions, whereas the rest (blue
rectangles) are the methods exposed as the public API. In
Figure~\ref{fig:romc_overview}, it can be easily observed that the
implementation follows Algorithm~\ref{alg:romc_algorithm}. The
training part includes all the steps until the computation of the
proposal regions, i.e., ~sampling the nuisance variables, defining the
optimisation problems, solving them, constructing the regions and
fitting local surrogate models. The inference part comprises of
evaluating the unnormalised posterior (and the normalised when is
possible), sampling and computing an expectation. We also provide some
utilities for inspecting the training process, such as plotting the
histogram of the final distances or visualising the constructed
bounding boxes. Finally, in the evaluation part, we provide two
methods for evaluating the inference; (a) computing the Effective
Sample Size (ESS) of the samples and (b) measuring the divergence
between the approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method,
  i.e., ABC Rejection Sampling), and we would like to measure the
  discrepancy between the two approximations.}

@@

\begin{Code}
elfi.ROMC(<output_node>, parallelize = True)
\end{Code}

@@

In the following code snippet, we code the model at \pkg{ELFI}:

\begin{Code}
import elfi
import scipy.stats as ss
import numpy as np

def simulator(t1, batch_size = 1, random_state = None):
    c = 0.5 - 0.5**4
    if t1 < -0.5:
        y = ss.norm(loc = -t1-c, scale = 1).rvs(random_state = random_state)
    elif t1 <= 0.5:
        y = ss.norm(loc = t1**4, scale = 1).rvs(random_state = random_state)
    else:
        y = ss.norm(loc = t1-c, scale = 1).rvs(random_state = random_state)
    return y

@@

# Elfi graph
t1 = elfi.Prior('uniform', -2.5, 5)
sim = elfi.Simulator(simulator, t1, observed = y)
d = elfi.Distance('euclidean', sim)

# Initialise the ROMC inference method
bounds = [(-2.5, 2.5)] # bounds of the prior
parallelize = True # activate parallel execution
romc = elfi.ROMC(d, bounds = bounds, parallelize = parallelize)
\end{Code}

### 2022-12-21 article/latex_files/ch4/4-1-training.tex
fixed spaces and commas

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.solve_problems(n1, use_bo=False, optimizer_args=None)
----------------------------------------------------------------------------
\end{Code}

\noindent
This method (a) draws the nuisance variables, (b) defines the
optimisation problems and (c) solves them using either a
gradient-based optimiser or the Bayesian optimisation (B0) scheme. The
three tasks are completed sequentially, as shown in
Figure~\ref{fig:romc_overview}. The definition of the optimisation
problems is performed by drawing \code{n1} integer numbers from a
discrete uniform distribution \(u_i \sim \mathcal{U}\{1,
2^{32}-1\}\). Each integer \(u_i\) is the seed used in \pkg{ELFI}'s
random simulator. The user may select the Bayesian Optimisation scheme
by setting the argument \code{use_bo=True} and pass custom arguments
to the optimizer through \code{optimize_args}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.distance_hist(**kwargs)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.estimate_regions(eps_filter, use_surrogate=None, fit_models=False)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.fit_posterior(args*)  # training in a single call
>>> romc.visualize_region(i)   # acceptance region
>>> romc.compute_eps(quantile) # estimates eps
----------------------------------------------------------------------------
\end{Code}

\noindent
The function \code{fit_posterior} is a syntactic sugar for applying
\code{solve_problems} and \\ \code{estimate_regions} into a single
step. The function \code{visualize_region} can be used for plotting
the bounding box around the optimal point, when the parameter space is
up to 2D. The argument \code{i} is the index of the corresponding
optimisation problem. Finally, \code{compute_eps} returns the
appropriate distance value \(d_{i=\kappa}^*\) where
\(\kappa = \lfloor quantile \cdot n \rfloor\) from the collection
\(\{ d_i^* \} \forall i = \{1, \ldots, n\}\) where \(n\) is the number
of accepted solutions. It can be used to automate the selection of the
threshold \(\epsilon\), e.g.\
\code{eps=romc.compute_eps(quantile=0.9)}.

@@

\begin{Code}
------------------------------ python snippet ------------------------------
  # Training (fitting) part
  n1 = 500 # number of optimisation problems
  seed = 21 # seed for solving the optimisation problems
  use_bo = False # set to True for switching to Bayesian optimisation

  # Training step-by-step
  romc.solve_problems(n1=n1, seed=seed, use_bo=use_bo)
  romc.theta_hist(bins=100) # plot hist to decide which eps to use

  eps = .75 # threshold for the bounding box 
  romc.estimate_regions(eps=eps) # build the bounding boxes

  romc.visualize_region(i=1) # for inspecting visually the bounding box

  # Equivalent one-line command
  # romc.fit_posterior(n1=n1, eps=eps, use_bo=use_bo, seed=seed)
----------------------------------------------------------------------------  
\end{Code}

>>>>

\begin{Code}
romc.solve_problems(n1, use_bo = False, optimizer_args = None)
\end{Code}

\noindent
This method (a) draws the nuisance variables, (b) defines the
optimisation problems and (c) solves them using either a
gradient-based optimiser or the Bayesian optimisation (B0) scheme. The
three tasks are completed sequentially, as shown in
Figure~\ref{fig:romc_overview}. The definition of the optimisation
problems is performed by drawing \code{n1} integer numbers from a
discrete uniform distribution \(u_i \sim \mathcal{U}\{1,
2^{32}-1\}\). Each integer \(u_i\) is the seed used in \pkg{ELFI}'s
random simulator. The user may select the Bayesian Optimisation scheme
by setting the argument \code{use_bo = True} and pass custom arguments
to the optimizer through \code{optimize_args}.

\begin{Code}
romc.distance_hist(**kwargs)
\end{Code}

@@

\begin{Code}
romc.estimate_regions(eps_filter, use_surrogate=None, fit_models=False)
\end{Code}

@@

\begin{Code}
romc.fit_posterior(args*)  # training in a single call
romc.visualize_region(i)   # acceptance region
romc.compute_eps(quantile) # estimates eps
----------------------------------------------------------------------------
\end{Code}

\noindent
The function \code{fit_posterior} is a syntactic sugar for applying
\code{solve_problems} and \\ \code{estimate_regions} into a single
step. The function \code{visualize_region} can be used for plotting
the bounding box around the optimal point, when the parameter space is
up to 2D. The argument \code{i} is the index of the corresponding
optimisation problem. Finally, \code{compute_eps} returns the
appropriate distance value \(d_{i=\kappa}^*\) where
\(\kappa = \lfloor quantile \cdot n \rfloor\) from the collection
\(\{ d_i^* \} \forall i = \{1, \ldots, n\}\) where \(n\) is the number
of accepted solutions. It can be used to automate the selection of the
threshold \(\epsilon\), e.g.,
\code{eps=romc.compute_eps(quantile=0.9)}.

@@

\begin{Code}
# Training (fitting) part
n1 = 500 # number of optimisation problems
seed = 21 # seed for solving the optimisation problems
use_bo = False # set to True for switching to Bayesian optimisation

# Training step-by-step
romc.solve_problems(n1 = n1, seed = seed, use_bo = use_bo)
romc.theta_hist(bins = 100) # plot hist to decide which eps to use

eps = .75 # threshold for the bounding box 
romc.estimate_regions(eps = eps) # build the bounding boxes

romc.visualize_region(i = 1) # for inspecting visually the bounding box

# Equivalent one-line command
# romc.fit_posterior(n1 = n1, eps = eps, use_bo = use_bo, seed = seed)
\end{Code}
